---
title: "Boolean Nested Effects Models \n
Inferring the logical signalling of pathways from indirect measurements
and perturbation biology."
author: "Martin Pirkl"
date: "`r Sys.Date()`"
graphics: yes
output: BiocStyle::pdf_document
vignette: >
%\VignetteIndexEntry{bnem}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---

## Introduction

Boolean Nested effects Models (B-NEM) are used to infer signalling pathways. In different experiments (conditions) members of a pathway (S-genes) are stimulated or inhibited, alone and in combination. In each experiment transcriptional targets (E-genes) of the pathway react differently and are higher or lower expressed depending on the condition. From these differential expression B-NEM infers Boolean functions presented as hyper-edges of a hyper-graph connecting parents and children in the pathway. For example if the signal is transducted by two parents A and B to a child C and the signal can be blocked with a knock-down of either one, they are connected by a typical AND-gate. If the signal is still transduced during a single knock-down, but blocked by the double knock-down of A and B, they activate C by an OR-gate. In general the state of child C is defined by a Boolean function
\[f: \left\{0,1\right\}^n \to \left\{0,1\right\},~C = f(A_1,\dots,A_n)\]
with its parents $A_i, i \in \left\{1,\dots,n\right\}$.

The B-NEM package is based on and uses many low level function of the CellNOptR package of Terfve et al., 2012.

## Installign and loading B-NEM

```{r, eval=FALSE}
install.packages("devtools")

library(devtools)

github_install(bnem)
```

```{r}
library(bnem)
library(mnem)
library(CellNOptR)
library(snowfall)
library(epiNEM)
library(matrixStats)

## load some pre-calculated results to speed up vignette building:
data(bnem_data)

greedy <- bnem_data$greedy
genetic <- bnem_data$genetic
greedy2 <- bnem_data$greedy2
greedy2b <- bnem_data$greedy2b

bcr_greedy <- bnem_data$bcr_greedy
bcr_genetic <- bnem_data$bcr_genetic

```

## Toy example for DAG

We show how to use B-NEM on a toy pathway presented by a directed
acyclic (hyper-)graph (DAG). B-NEM demands several objects as
input. The two main objects are the differential gene expression
(data) and prior knowledge respectively the search space.

First we create a prior knowledge network (PKN). The PKN will have two
S-genes without parents. We define these as stimuli, which means they
are set to $1$ in experiments, where they are stimulated and $0$
otherwise. All other down-stream S-genes have the potential to be
inhibited $(0)$. If they are not inhibited their state is calculated
according to their parents' state and a given Boolean function.

```{r}
set.seed(2579)
## alternative seed and also a great song:
## set.seed(9247)

## to get the while loop started,
## which makes sure we get a PKN with exactly two stimuli (or what the amount you want):
stimuli <- "dummy"

while(length(stimuli) != 2) {

    ## random Boolean graph without cycles,
    ## maximal 25 edges, maximal edge size of 1 (normal graph) and maximal 10 S-genes:
    dnf <- randomDnf(10, max.edges = 25, max.edge.size = 1, dag = TRUE)

    ## all S-genes:
    cues <- sort(unique(gsub("!", "", unlist(strsplit(unlist(strsplit(dnf, "=")), "\\+")))))

    ## parents:
    inputs <- unique(unlist(strsplit(gsub("!", "", gsub("=.*", "", dnf)), "=")))

    ## children:
    outputs <- unique(gsub(".*=", "", dnf))

    ## parents which are no children are stimuli:
    stimuli <- inputs[which(!(inputs %in% outputs))]
    
}

inhibitors <- unique(c(inputs, outputs))
## S-genes which are no stimuli are inhibitors
inhibitors <- inhibitors[-which(inhibitors %in% stimuli)]
```

The following figure shows the PKN. Red "tee" arrows depict repression the others activation of the child. The stimulated S-genes are diamond shaped.

```{r, fig.width = 5, fig.height = 5}
plotDnf(dnf, stimuli = stimuli)
```

In the next step we convert the dnf object into a PKN object and
extend it to a Boolean hyper-graph, which represents an apriori restricted search space of Boolean networks.

```{r}
sifMatrix <- NULL

for (i in dnf) {
    inputs2 <- unique(unlist(strsplit(gsub("=.*", "", i), "=")))
    output <- unique(gsub(".*=", "", i))
    for (j in inputs2) {
        j2 <- gsub("!", "", j)
        if (j %in% j2) {
            sifMatrix <- rbind(sifMatrix, c(j, 1, output))
        } else {
            sifMatrix <- rbind(sifMatrix, c(j2, -1, output))
        }
    }
}
write.table(sifMatrix, file = "temp.sif", sep = "\t",
            row.names = FALSE, col.names = FALSE, quote = FALSE)
PKN <- readSIF("temp.sif")
## unlink("temp.sif")

## create metainformation (which S-genes are perturbed in which experiments):
CNOlist <- dummyCNOlist(stimuli = stimuli, inhibitors = inhibitors,
                        maxStim = 2, maxInhibit = 1, signals = NULL)

## extend the model:
model <- preprocessing(CNOlist, PKN, maxInputsPerGate=100, verbose = TRUE)
```

We suggest to take a look at the sif file. In future analyses it is easier to just provide a suitable sif file for the investigated pathway.

In an real application the underlaying ground truth network (GTN) is
not known. However in our toy example we define one.

```{r}
## define a bitstring denoting the present and absent edges:
bString <- absorption(sample(c(0,1), length(model$reacID), replace = TRUE), model)

## simulate S-gene states for all possible conditions:
steadyState <- steadyState2 <- simulateStatesRecursive(CNOlist, model, bString)

## we find constitutively active S-genes with the following
## (they are boring, so we avoid them):
ind <- grep(paste(inhibitors, collapse = "|"), colnames(steadyState2))
steadyState2[, ind] <- steadyState2[, ind] + CNOlist@inhibitors

## this while loop makes sure we get a gtn,
## which actually affects all vertices
## and no vertices are constitutively active:
while(any(apply(steadyState, 2, sd) == 0) | any(apply(steadyState2, 2, sd) == 0)) {

    bString <- absorption(sample(c(0,1), length(model$reacID), replace = TRUE), model)

    steadyState <- steadyState2 <- simulateStatesRecursive(CNOlist, model, bString)

    ind <- grep(paste(inhibitors, collapse = "|"), colnames(steadyState2))
    steadyState2[, ind] <- steadyState2[, ind] + CNOlist@inhibitors

}
```

The following figure shows our random GTN, which is a subset of the
hyper-graph representing the apriori restricted search space.

```{r, fig.width = 5, fig.height = 5}
plotDnf(model$reacID[as.logical(bString)], stimuli = stimuli)
```

We use this GTN to simulate data.

```{r}
## the expression data with 10 E-genes for each S-gene and 3 replicates:
exprs <- t(steadyState)[rep(1:ncol(steadyState), 10), rep(1:nrow(steadyState), 3)]

## we calculate the foldchanges (expected S-gene response scheme, ERS)
## between certain condtions (e.g. control vs stimulation):
ERS <- computeFc(CNOlist, t(steadyState))

                                        # the next step reduces the ERS to a sensible set of comparisons;
## e.g. we do not want to compare stimuli vs inhibition,
## but stimuli vs (stimuli,inhibition):
stimcomb <- apply(expand.grid(stimuli, stimuli), c(1,2), as.character)
stimuli.pairs <- apply(stimcomb, 1, paste, collapse = "_")

## this is the usual setup,
## but "computeFc" calculates a lot more contrasts, which can also be used, if preferred:
ind <- grep(paste(c(paste("Ctrl_vs_", c(stimuli, inhibitors), sep = ""),
                    paste(stimuli, "_vs_", stimuli, "_",
                          rep(inhibitors, each = length(stimuli)), sep = ""),
                    paste(stimuli.pairs, "_vs_", stimuli.pairs, "_",
                          rep(inhibitors, each = length(stimuli.pairs)), sep = "")),
                  collapse = "|"), colnames(ERS))
ERS <- ERS[, ind]

## same as before with the expression values, we have 10 E-genes each and 3 replicates:
fc <- ERS[rep(1:nrow(ERS), 10), rep(1:ncol(ERS), 3)]

## we add some Gaussian noise:
fc <- fc + rnorm(length(fc), 0, 1)

## in real applications some E-genes are negatively regulated,
## hence we flip some foldchanges:
flip <- sample(1:nrow(fc), floor(0.33*row(fc)))
fc[flip, ] <- fc[flip, ]*(-1)

## don't forget to set rownames (usually gene symbols, ensemble ids, entrez ids, ...)
rownames(fc) <- paste(rownames(fc), 1:nrow(fc), sep = "_")
print(fc[1:6, c(1:3,(ncol(fc)-2):ncol(fc))])
```

B-NEM uses differential expression between experiments to infer the
pathway logics. For example look at the colnames of fc (=foldchanges
of E-genes (rows)) and remember that `r stimuli` are our
stimulated S-genes and the rest possibly inhibited. Thus in the first
column of fc we have the contrast `r gsub(".*_", "", colnames(fc)[1])` $-$ control. In the control no S-genes are perturbed.

We search for the GTN in our restricted network space. Each network is a sub-graph of the full hyper-graph model\$reacID. We initialise the search with a starting network and greedily search the neighbourhood.

```{r}
## we start with the empty graph:
initBstring <- reduceGraph(rep(0, length(model$reacID)), model, CNOlist)
## or a fully connected graph:
## initBstring <- reduceGraph(rep(1, length(model$reacID)), model, CNOlist)

## paralellize for several threads on one machine or multiple machines
## see package "snowfall" for details
parallel <- 2 # NULL for serialization
## or distribute to 30 threads on four different machines:
## parallel <- list(c(4,16,8,2), c("machine1", "machine2", "machine3", "machine4"))
```
```{r, eval = TRUE}
## greedy search:
greedy <- bnem(search = "greedy",
               fc=fc,
               exprs=exprs, # not used, if fc is defined
               CNOlist=CNOlist,
               model=model,
               parallel=parallel,
               initBstring=bString*0,#initBstring,
               draw = FALSE, # TRUE: draw network at each step
               verbose = TRUE, # TRUE: print changed (hyper-)edges and score improvement
               maxSteps = Inf,
               method = "s"
               )

greedy2 <- bnem(search = "greedy",
                fc=fc,
                exprs=exprs, # not used, if fc is defined
                CNOlist=CNOlist,
                model=model,
                parallel=parallel,
                initBstring=bString*0,#initBstring,
                draw = FALSE, # TRUE: draw network at each step
                verbose = TRUE, # TRUE: print changed (hyper-)edges and score improvement
                maxSteps = Inf,
                method = "llr"
                )

print(greedy$scores)
print(greedy2$scores)

resString <- greedy2$bString

print(sum(bString == 1 & resString == 1)/
      (sum(bString == 1 & resString == 1) + sum(bString == 1 & resString == 0)))
print(sum(bString == 0 & resString == 0)/
      (sum(bString == 0 & resString == 0) + sum(bString == 0 & resString == 1)))

```

We take a look at the efficiency of the search algorithm with
sensitivity and specificity of the hyper-eges for the optimized network and the accuracy of its ERS (similar to the truth table). Since several networks produce the same ERS, the learned hyper-graph can differ from the GTN and still be $100\%$ accurate.

```{r, fig.width = 10, fig.height = 5}
par(mfrow=c(1,2))

## GTN:
plotDnf(model$reacID[as.logical(bString)], main = "GTN", stimuli = stimuli)

## greedy optimum:
plotDnf(model$reacID[as.logical(resString)], main = "greedy optimum", stimuli = stimuli)
```

```{r}
## hyper-edge sensitivity and specificity:
print(sum(bString == 1 & resString == 1)/
      (sum(bString == 1 & resString == 1) + sum(bString == 1 & resString == 0)))
print(sum(bString == 0 & resString == 0)/
      (sum(bString == 0 & resString == 0) + sum(bString == 0 & resString == 1)))

## accuracy of the expected response scheme (can be high even, if the networks differ):
ERS.res <- computeFc(CNOlist, t(simulateStatesRecursive(CNOlist, model, resString)))
ERS.res <- ERS.res[, which(colnames(ERS.res) %in% colnames(ERS))]
print(sum(ERS.res == ERS)/length(ERS))
```

In our seeded example 2574 the optimum network is indeed the GTN. If
we set the alternative seed at the beginning, neither the network nor
the correct ERS is found, if we initialise the search with the empty
network. If we start with the fully connected normal graph, the ERS is
resolved to a $100$\%, while the greedy optimum still differs from the
GTN. Thus the GTN and the greedy optimum are equivalent, but the GTN
is larger (more outgoing edges) and thus not preferred.

After optimization we look at the data and how well the greedy optimum
explains the E-genes. The lower the score the better the fit.

```{r, fig.width = 10, fig.height = 10}
fitinfo <- validateGraph(CNOlist, fc=fc, exprs=exprs, model = model, bString = resString,
                         Sgene = 1, Egenes = 1000, cexRow = 0.8, cexCol = 0.7, xrot = 45,
                         Colv = TRUE, Rowv = TRUE, dendrogram = "both", bordercol = "lightgrey",
                         aspect = "iso", sub = "")
```

The bottom row shows the ERS of S-gene
```r colnames(CNOlist@signals[[1]])[1]``` and the other rows show the
observed response scheme (ORS) of the
```r colnames(CNOlist@signals[[1]])[1]```-regulated E-genes. Even
though the Gaussian noise makes the data look almost random, the
greedy optimum is equal to the GTN.
Alternatively to the greedy neighbourhood search a genetic algorithm
and exhaustive search are available. The exhaustive search is not recommended for search spaces with more than 20 hyper-edges.

```{r, eval = FALSE}
## genetic algorithm:
genetic <- bnem(search = "genetic",
           fc=fc,
           exprs=exprs,
           parallel = parallel,
           CNOlist=CNOlist,
           model=model,
           initBstring=initBstring,
           popSize = 10,
           stallGenMax = 10,
           draw = FALSE,
           verbose = FALSE
           )
```
```{r}
resString <- genetic$bString
```

```{r, eval = FALSE}
## exhaustive search:
exhaustive <- bnem(search = "exhaustive",
                   parallel = parallel,
                   CNOlist=CNOlist,
                   fc=fc,
                   exprs=exprs,
                   model=model
                   )

resString <- exRun$bString
```

## Stimulated and inhibited S-genes can overlap

In this section we show how to use B-NEM, if stimuli and inhibitors
overlap. Additionally we want to show that B-NEM can resolve cycles. For this we allow the PKN to have cycles, but no repression, because repression can lead to an unresolvable ERS. See Pirkl et al., 2016 for details.

```{r}
## we do not force a DAG but do not allow repression:
dnf <- randomDnf(10, max.edges = 25, max.edge.size = 1, dag = FALSE, negation = FALSE)
cues <- sort(unique(gsub("!", "", unlist(strsplit(unlist(strsplit(dnf, "=")), "\\+")))))
inputs <- unique(unlist(strsplit(gsub("!", "", gsub("=.*", "", dnf)), "=")))
outputs <- unique(gsub(".*=", "", dnf))
stimuli <- cues # c(inputs[which(!(inputs %in% outputs))], cues[sample(1:length(cues), 2)])
inhibitors <- cues
```

We look for stimuli which are also inhibited. For those we add additional stimuli S-genes. The stimuli S-gene (parent) and the inhibitor S-gene (child) are connected by a positive edge.

```{r}
both <- stimuli[which(stimuli %in% inhibitors)]
for (i in both) {
    dnf <- gsub(i, paste(i, "inhibit", sep = ""), dnf)
    dnf <- c(dnf, paste(i, "stim=", i, "inhibit", sep = ""))
    stimuli <- gsub(i, paste(i, "stim", sep = ""), stimuli)
    inhibitors  <- gsub(i, paste(i, "inhibit", sep = ""), inhibitors)
}
```

The next figure shows the cyclic PKN with extra stimuli S-genes. Notice, this way the inhibition of the S-genes overrules the stimulation.

```{r, fig.width = 5, fig.height = 9}
plotDnf(dnf, stimuli = stimuli)
```

Similar to before we create the full network search space, draw a GTN, simulate data and search for the optimal network.

```{r, fig.width = 10, fig.height = 9}
sifMatrix <- NULL
for (i in dnf) {
    inputs2 <- unique(unlist(strsplit(gsub("=.*", "", i), "=")))
    output <- unique(gsub(".*=", "", i))
    for (j in inputs2) {
        j2 <- gsub("!", "", j)
        if (j %in% j2) {
            sifMatrix <- rbind(sifMatrix, c(j, 1, output))
        } else {
            sifMatrix <- rbind(sifMatrix, c(j2, -1, output))
        }
    }
}
write.table(sifMatrix, file = "temp.sif", sep = "\t", row.names = FALSE, col.names = FALSE,
            quote = FALSE)
PKN <- readSIF("temp.sif")
## unlink("temp.sif")
CNOlist <- dummyCNOlist(stimuli = stimuli, inhibitors = inhibitors,
                        maxStim = 2, maxInhibit = 1, signals = NULL)
model <- preprocessing(CNOlist, PKN, maxInputsPerGate=100, verbose = FALSE)

bString <- absorption(sample(c(0,1), length(model$reacID), replace = TRUE), model)
steadyState <- steadyState2 <- simulateStatesRecursive(CNOlist, model, bString)
ind <- grep(paste(inhibitors, collapse = "|"), colnames(steadyState2))
steadyState2[, ind] <- steadyState2[, ind] + CNOlist@inhibitors
while(any(apply(steadyState, 2, sd) == 0) | any(apply(steadyState2, 2, sd) == 0)) {
    bString <- absorption(sample(c(0,1), length(model$reacID), replace = TRUE), model)
    steadyState <- steadyState2 <- simulateStatesRecursive(CNOlist, model, bString)
    ind <- grep(paste(inhibitors, collapse = "|"), colnames(steadyState2))
    steadyState2[, ind] <- steadyState2[, ind] + CNOlist@inhibitors
}
## we make sure the stimulations work:
bString[grep("stim", model$reacID)] <- 1
bString <- absorption(bString,model)

plotDnf(model$reacID[as.logical(bString)], stimuli = stimuli)

exprs <- t(steadyState)[rep(1:ncol(steadyState), 10), rep(1:nrow(steadyState), 3)]
ERS <- computeFc(CNOlist, t(steadyState))
stmcomb <- apply(expand.grid(stimuli, stimuli), c(1,2), as.character)
stimuli.pairs <- apply(stimcomb, 1, paste, collapse = "_")
ind <- grep(paste(c(paste("Ctrl_vs_", c(stimuli, inhibitors), sep = ""),
                    paste(stimuli, "_vs_", stimuli, "_",
                          rep(inhibitors, each = length(stimuli)), sep = ""),
                    paste(stimuli.pairs, "_vs_", stimuli.pairs, "_",
                          rep(inhibitors, each = length(stimuli.pairs)), sep = "")),
                  collapse = "|"), colnames(ERS))
ERS <- ERS[, ind]
fc <- ERS[rep(1:nrow(ERS), 10), rep(1:ncol(ERS), 3)]
fc <- fc + rnorm(length(fc), 0, 1)
flip <- sample(1:nrow(fc), floor(0.33*row(fc)))
fc[flip, ] <- fc[flip, ]*(-1)
rownames(fc) <- paste(rownames(fc), 1:nrow(fc), sep = "_")
print(fc[1:6, c(1:3,(ncol(fc)-2):ncol(fc))])

initBstring <- reduceGraph(rep(0, length(model$reacID)), model, CNOlist)
```
```{r, eval = FALSE}
greedy2 <- bnem(search = "greedy",
                CNOlist=CNOlist,
                fc=fc,
                exprs=exprs,
                model=model,
                parallel=parallel,
                initBstring=initBstring,
                draw = FALSE,
                verbose = FALSE,
                maxSteps = Inf,
                method = "llr"
                )
```
```{r, fig.width = 10, fig.height = 9}
resString2 <- greedy2$bString

par(mfrow=c(1,2))
plotDnf(model$reacID[as.logical(bString)], main = "GTN", stimuli = stimuli)
plotDnf(model$reacID[as.logical(resString2)], main = "greedy optimum", stimuli = stimuli)

print(sum(bString == 1 & resString2 == 1)/
      (sum(bString == 1 & resString2 == 1) + sum(bString == 1 & resString2 == 0)))
print(sum(bString == 0 & resString2 == 0)/
      (sum(bString == 0 & resString2 == 0) + sum(bString == 0 & resString2 == 1)))
ERS.res <- computeFc(CNOlist, t(simulateStatesRecursive(CNOlist, model, resString2)))
ERS.res <- ERS.res[, which(colnames(ERS.res) %in% colnames(ERS))]
print(sum(ERS.res == ERS)/length(ERS))
```

The greedy optimum looks different from the GTN, even though they share a
lot, but not all edges (reduced sensitivity and specificity). However the accuracy of the ERS is still $100\%$.

## Pre-attach E-genes
One additional challenge for B-NEM compared to methods like CellNetOptimizer is the fact, that B-NEM optimizes the signalling pathway and simulataneously the attachment of the E-genes. However, it is possible to include prior knowledge into the search.

We just have to create a list object, which holds the ERS and prior information about the E-genes.

```{r}
egenes <- list()

for (i in cues) {
    egenes[[i]] <- rownames(fc)[grep(i, rownames(fc))]
}
```
```{r, eval = FALSE}
initBstring <- reduceGraph(rep(0, length(model$reacID)), model, CNOlist)
greedy2b <- bnem(search = "greedy",
                 CNOlist=CNOlist,
                 fc=fc,
                 exprs=exprs,
                 egenes=egenes,
                 model=model,
                 parallel=parallel,
                 initBstring=initBstring,
                 draw = FALSE,
                 verbose = FALSE,
                 maxSteps = Inf,
                 method = "llr"
                 )
```
```{r}
resString3 <- greedy2b$bString
```

We attach every E-gene to its real parent in the for loop. If an E-gene is only included once in the egenes object, it's position is not learned, but fixed during the optimization of the signalling pathway. Alternatively, we can include one E-gene several times for just a subset of S-genes. This way S-genes, which do not have the E-genes included in their E-gene set are excluded as potential parents.

```{r}
print(sum(bString == 1 & resString3 == 1)/
      (sum(bString == 1 & resString3 == 1) + sum(bString == 1 & resString3 == 0)))
print(sum(bString == 0 & resString3 == 0)/
      (sum(bString == 0 & resString3 == 0) + sum(bString == 0 & resString3 == 1)))
ERS.res <- computeFc(CNOlist, t(simulateStatesRecursive(CNOlist, model, resString3)))
ERS.res <- ERS.res[, which(colnames(ERS.res) %in% colnames(ERS))]
print(sum(ERS.res == ERS)/length(ERS))
```

In our toy example, fixing the correct E-genes to their parents increases the accuracy of the network and even resolves the ERS to 100 percent.

## Visualizing network residuals
We can also quantify how well the attached E-genes fit to the learned network. See the references for more details.

```{r, eval = FALSE}
residuals <- findResiduals(resString3, CNOlist, model, fc, verbose = FALSE)
## verbose = TRUE plots the residuals matrices
```

Row denote S-genes in the network. Columns denote Contrasts between two experiments. Green colors in the left matrix show the score improves, if no (0) or a negative (-1) response in the network's ERS is changed to positive (+1). Red colors show a zero changed to positive. The right matrix shows the same for switched +1 and -1.

## B-Cell receptor signalling

In this section we analyze the B-Cell receptor (BCR) signalling
data. The dataset consists of one stimulated S-gene (BCR), three
S-genes with available single inhibitions (Tak1, Pik3, Erk) and three S-genes
with up to triple inhibitions.

```{r}
data(bcr)
head(fc)
```

We build a PKN to incorporate biological knowledge and account for missing combinatorial inhibitions.

```{r}
negation <- FALSE # what happens if we allow negation?
sifMatrix <- numeric()
for (i in "BCR") {
    sifMatrix <- rbind(sifMatrix, c(i, 1, c("Pi3k")))
    sifMatrix <- rbind(sifMatrix, c(i, 1, c("Tak1")))
    if (negation) {
        sifMatrix <- rbind(sifMatrix, c(i, -1, c("Pi3k")))
        sifMatrix <- rbind(sifMatrix, c(i, -1, c("Tak1")))
    }
}
for (i in c("Pi3k", "Tak1")) {
    for (j in c("Ikk2", "p38", "Jnk", "Erk", "Tak1", "Pi3k")) {
        if (i %in% j) { next() }
        sifMatrix <- rbind(sifMatrix, c(i, 1, j))
        if (negation) {
            sifMatrix <- rbind(sifMatrix, c(i, -1, j))
        }
    }
}
for (i in c("Ikk2", "p38", "Jnk")) {
    for (j in c("Ikk2", "p38", "Jnk")) {
        if (i %in% j) { next() }
        sifMatrix <- rbind(sifMatrix, c(i, 1, j))
        if (negation) {
            sifMatrix <- rbind(sifMatrix, c(i, -1, j))
        }
    }
}

write.table(sifMatrix, file = "temp.sif", sep = "\t",
            row.names = FALSE, col.names = FALSE, quote = FALSE)
PKN <- readSIF("temp.sif")
unlink("temp.sif")
```

In the next step, we create the meta information. This ensures, that
we simulate all the conditions, which are actually available in the
data. Furthermore we build our boolean search space based on the PKN.

```{r}
CNOlist <- dummyCNOlist(stimuli = "BCR",
                        inhibitors = c("Tak1", "Pi3k", "Ikk2", "Jnk", "p38", "Erk"),
                        maxStim = 1, maxInhibit = 3)

model <- preprocessing(CNOlist, PKN)
```

In the final step we learn the network with the genetic algorithm and deterministic greedy search.

```{r, fig.width = 10, fig.height = 5}
initBstring <- rep(0, length(model$reacID))
```
```{r, eval = FALSE}
bcr_genetic <- bnem(search = "genetic",
                    fc=fc,
                    CNOlist=CNOlist,
                    model=model,
                    parallel=2,
                    initBstring=initBstring,
                    draw = FALSE,
                    verbose = FALSE
                    )
```
```{r}
print(min(bcr_genetic$scores))

initBstring <- rep(0, length(model$reacID))
```
```{r, eval = FALSE}
bcr_greedy <- bnem(search = "greedy",
                   fc=fc,
                   CNOlist=CNOlist,
                   model=model,
                   parallel=2,
                   initBstring=initBstring,
                   draw = FALSE,
                   verbose = FALSE
                   )
```
```{r, fig.width = 10, fig.height = 5}
print(min(bcr_greedy$scores[[1]]))

par(mfrow=c(1,2))
plotDnf(PKN$reacID, main = "PKN", stimuli = "BCR")
plotDnf(bcr_genetic$graph, main = "genetic optimum", stimuli = "BCR")
plotDnf(bcr_greedy$graph, main = "greedy optimum", stimuli = "BCR")
```

```{r}
sessionInfo()
```

*__References:__*

Pirkl, Martin, Hand, Elisabeth, Kube, Dieter, & Spang, Rainer. 2016. Analyzing synergistic and non-synergistic interactions in signalling pathways using Boolean Nested Effect Models. \textit{Bioinformatics}, 32(6), 893–900.

Pirkl, Martin. 2016. Indirect inference of synergistic and
alternative signalling of intracellular pathways. University of
Regensburg.

Saez-Rodriguez, Julio, Alexopoulos, Leonidas G, Epperlein, Jonathan, Samaga, Regina, Lauffenburger, Douglas A, Klamt, Steffen, & Sorger, Peter K. 2009. Discrete logic modelling as a means to link protein signalling networks with functional analysis of mammalian signal transduction. Mol Syst Biol, 5, 331.\\

C Terfve, T Cokelaer, A MacNamara, D Henriques, E Goncalves, MK Morris, M van Iersel, DA Lauffenburger, J Saez-Rodriguez. CellNOptR: a flexible toolkit to train protein signaling networks to data using multiple logic formalisms. BMC Systems Biology, 2012, 6:133.
